---
title: "Bulk-RNAseq"
author: "Ting Zhai"
date: "2024-12-12"
output: html_document
---

For this group project, we use bulk RNA-seq data from GEO: GSE174367, to provide real-world application of our deconvolution approach. 

# Bulk data - initial inspection

## Read in bulk data 

```{r}
load("~/Downloads/GSE174367_bulkRNA_processed.rda")

str(normExpr.reg)
str(targets)

class(targets)

# assign gene names
library(biomaRt)
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
genes <- rownames(normExpr.reg)
gene_info <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"), filters = "ensembl_gene_id", values = genes, mart = ensembl)

normExpr.reg$hugo_name <- gene_info$hgnc_symbol[match(rownames(normExpr.reg), gene_info$ensembl_gene_id)]

# Check for missing HUGO names
sum(is.na(normExpr.reg$hugo_name) | normExpr.reg$hugo_name == "")
# Check for duplicates in HUGO names
duplicated_hugo <- normExpr.reg$hugo_name[duplicated(normExpr.reg$hugo_name)]
print(duplicated_hugo)  # List of duplicate HUGO names
# clean up the expression matrix
normExpr.reg$hugo_name[normExpr.reg$hugo_name == "" | is.na(normExpr.reg$hugo_name)] <- rownames(normExpr.reg)
```

## Umap visualization 

```{r}
load("./data/GSE174367_bulkRNA_processed.rda")
sum(is.na(normExpr.reg))
norm_counts_t <- t(normExpr.reg)
# Load libraries
library(umap)
library(ggplot2)

# Perform UMAP
umap_result <- umap(norm_counts_t)

# Create a dataframe for plotting
umap_df <- data.frame(UMAP1 = umap_result$layout[, 1],
                      UMAP2 = umap_result$layout[, 2],
                      Condition = targets$Diagnosis)  # Replace with your group variable

# UMAP Plot
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, fill = Condition)) +
  geom_point(size = 3.6, alpha = 0.6, shape=21) + scale_fill_viridis(discrete = TRUE, option="C") +
  labs(title = "",
       x = "UMAP 1",
       y = "UMAP 2") +
  theme_classic() 
```


# Data pre-processing 

First we need to clean up the data and prepare it for the deconvolution analysis. This includes assign gene names to replace Ensembl IDs, remove missing or empty gene names, and subset the data to include only the genes present in the reference columns. For the deconvolution, we will also split the data into case and control groups, to improve the accuracy of the deconvolution analysis.

```{r preprocessing}
# set work directory 
setwd("~/Documents/R_working/bst228")

# Load necessary libraries
library(biomaRt)

# Load bulk data
load("./data/GSE174367_bulkRNA_processed.rda")

# Inspect the structure of loaded objects
str(normExpr.reg)
str(targets)
class(targets)

# Set up Ensembl biomart connection to get gene annotations
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
genes <- rownames(normExpr.reg)
gene_info <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"), 
                   filters = "ensembl_gene_id", 
                   values = genes, 
                   mart = ensembl)

# Add HUGO gene symbols to normExpr.reg
normExpr.reg$hugo_name <- gene_info$hgnc_symbol[match(rownames(normExpr.reg), gene_info$ensembl_gene_id)]

# Replace missing or empty HUGO symbols with their Ensembl IDs
missing_hugo <- is.na(normExpr.reg$hugo_name) | normExpr.reg$hugo_name == ""
normExpr.reg$hugo_name[missing_hugo] <- rownames(normExpr.reg)[missing_hugo]

# (Optional) Check for duplicates in HUGO names, just to be aware
duplicated_hugo <- normExpr.reg$hugo_name[duplicated(normExpr.reg$hugo_name)]
if (length(duplicated_hugo) > 0) {
  message("Warning: There are duplicated HUGO names.")
  print(duplicated_hugo)
  # Consider how to handle duplicates if necessary (e.g., aggregating counts)
}

# Load the reference file
ref <- read.csv("./data/reference.csv", row.names = 1)
# list of reference ids
ad_ids <- c("AD00802", "AD00803", "AD01103", "AD01303", 
               "AD01304", "AD04902", "AD04904", "AD05701", "AD06302")
control_ids <- c("AD00801", "AD01101", "AD01102", "AD01301", "AD03203", 
                 "AD04901", "AD04903", "AD05703", "AD05704", "AD06301")

# Subset bulk data (Exp) to include only the genes present in the reference columns
Exp <- normExpr.reg[normExpr.reg$hugo_name %in% colnames(ref), ]
rownames(Exp) <- Exp$hugo_name
Exp <- Exp[ , -ncol(Exp)]  # remove the hugo_name column
Exp <- round(as.matrix(Exp))
Exp[Exp < 0] <- 0

# Subset ref data to include only genes present in Exp
ref <- ref[ , rownames(Exp)]

## at this point, save cleaned data for future use 
saveRDS(Exp, file = "./data/Exp_cleaned.rds")
saveRDS(ref, file = "./data/ref_cleaned.rds")
saveRDS(targets, file = "./data/targets.rds")

## save case and control specific data to directly call in the deconvolution script
Exp_case <- Exp[ , targets$SampleID[targets$Diagnosis == "AD"]]
Exp_control <- Exp[ , targets$SampleID[targets$Diagnosis == "Control"]]
ref_case <- ref[grepl(paste0("_(", paste(ad_ids, collapse = "|"), ")$"), rownames(ref)), ]
ref_control <- ref[grepl(paste0("_(", paste(control_ids, collapse = "|"), ")$"), rownames(ref)), ]

saveRDS(Exp_case, file = "./data/Exp_case.rds")
saveRDS(Exp_control, file = "./data/Exp_control.rds")
saveRDS(ref_case, file = "./data/ref_case.rds")
saveRDS(ref_control, file = "./data/ref_control.rds")
```


# Deconvolution of real bulk

```{bash terminal, eval=FALSE}
# This script trunk should be run in terminal to fully utilize the parallel processing capabilities of the Gibbs sampler as defined in the source script.

# Runtime on a local machine may vary depending on the number of cores available: 10 cores 20 GB RAM give around 3 hours for each. 

# Run the deconvolution script for case samples
./scripts/decon_case.R

# Run the deconvolution script for control samples
./scripts/decon_control.R 
```


# Post-processing analysis 

```{r}
# Load the output from the Gibbs sampler
res_case <- readRDS("./output/bulk_case_output.rds")
res_control <- readRDS("./output/bulk_control_output.rds")
Exp = readRDS("./data/Exp_cleaned.rds")
ref = readRDS("./data/ref_cleaned.rds")
Exp_case = readRDS("./data/Exp_case.rds")
Exp_control = readRDS("./data/Exp_control.rds")
```


## MCMC diagnostics 

```{r}
source("./scripts/diagnostics.R") # Load the diagnostics functions

# Get MCMC diagnostic metrics (rhat, ess) for the case and control samples
diagnostics_case <- mcmc_diagnostics_metrics(res_case$theta_n, res_case$theta_n_samples, threshold = 1e-8)

diagnostics_control <- mcmc_diagnostics_metrics(res_control$theta_n, res_control$theta_n_samples, threshold = 1e-8)

# Summarize the diagnostics
 hist(sapply(diagnostics_case, function(x) x$rhat))
 summary(sapply(diagnostics_case, function(x) x$effective_size))
head(diagnostics_case)
head(diagnostics_control)

# Plot the MCMC diagnostics for a specific sample and cell type
plot_mcmc_diagnostics(res = res_case,  
                      sample_name = "Sample-26", 
                      cell_type_name = "Astrocytes_AD00803",
                      sample_names = colnames(Exp_case),
                      cell_type_names = rownames(ref_case))

# Plot the diagnostics metrics obtained earlier 
dense_df <- data.frame(rhat = sapply(diagnostics_case, function(x) x$rhat),
                       ess = sapply(diagnostics_case, function(x) x$effective_size))

p<- ggplot(dense_df, aes(x = effective_size)) + 
  geom_density(fill = "lightblue", color = "grey", alpha = 0.5) + 
  xlab("Effective sample size") + ylab("Density") + theme_minimal()
ggsave("./output/ess_density.png", plot = p, width = 4, height = 2, units = "in")

p<- ggplot(dense_df, aes(x = rhat)) + 
  geom_density(fill = "lightblue", color = "grey", alpha = 0.5) + 
  xlab("Rhat") + ylab("Density") + theme_minimal()
ggsave("./output/rhat_density.png", plot = p, width = 4, height = 2, units = "in")

```


## Proportion calculation 

```{r}
# obtain cell type proportions from deconvolution results
prop_case <- as.data.frame(res_case$theta_n)
prop_control <- as.data.frame(res_control$theta_n)

# Combine case and control data
prop_all <- bind_rows(process_sample_proportions(prop_case) 
                     , process_sample_proportions(prop_control)
                   )
# Calculate summed proportions for each cell type across references, grouped by sample
n_chains = 5

prop_summed <- prop_all %>%
  group_by(sample, cell_type) %>%
  summarise(total_proportion = sum(proportion, na.rm = TRUE)/n_chains, .groups = "drop") %>% 
  pivot_wider(names_from = cell_type, values_from = total_proportion)

# View the processed proportions
head(prop_summed)

# merge with patient data
dat <- targets %>% 
  select(SampleID, Neuropath.Dx.1, Age, Sex, PMI, APoE, Tangle.Stage, Diagnosis, Clinical.Syndrome) %>%
  left_join(prop_summed, by = c("SampleID" = "sample"))
```


## Visualization 

```{r}
library(ggplot2)
library(viridis)
# Plot the celltype proportion variability by reference
ggplot(prop_all, aes(x = cell_type, y = proportion, fill = reference)) +
  geom_bar(stat = "identity") + scale_fill_viridis(discrete = TRUE) +
  labs(
    title = "Stacked Bar Plot of Reference Contribution by Cell Type",
    x = "Cell Type",
    y = "Proportion",
    fill = "Reference"
  ) +
  theme_minimal()

ggplot(prop_all, aes(x = cell_type, y = proportion, fill = reference)) +
  geom_bar(stat = "identity") + scale_fill_viridis(discrete = TRUE) +
  labs(title = "Stacked Bar Plot of Reference Contribution by Cell Type",
       x = "",
       y = "Deconvoluted Proportion") +
  theme(
    legend.text = element_text(size = 8),       # Shrink legend text size
    legend.title = element_text(size = 10),     # Shrink legend title size
    legend.key.size = unit(0.3, "cm"),          # Reduce legend key size
    legend.spacing.y = unit(0.1, "cm")          # Add spacing between legend items
  ) +
  guides(fill = guide_legend(byrow = TRUE)) 
```


```{r}
# make long format for plotting
dat_long <- dat %>% 
  pivot_longer(cols = c(Astrocytes, `Excitatory neurons`, `Inhibitory neurons`), names_to = "CellType", values_to = "Proportion")
# cell type by case vs control in patient dat
ggplot(dat_long, aes(x = CellType, y = Proportion, color = Diagnosis)) +
  geom_jitter(alpha=0.6) + scale_color_viridis(discrete = TRUE) +
  labs(
    title = "Cell Proportions by Diagnosis",
    x = "Cell Type",
    y = "Deconvoluted Proportion",
    color = "Diagnosis"
  ) +
  theme_minimal()
ggplot(dat_long, aes(x = Diagnosis, y = Proportion, color = Diagnosis, fill = Diagnosis)) + scale_fill_viridis(discrete = TRUE) + scale_color_viridis(discrete = TRUE) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(alpha = 0.6, size = 1.5, width = 0.2) +
  facet_wrap(~CellType) +
    labs(
    title = "Cell Proportions by Diagnosis",
    x = "",
    y = "Deconvoluted Proportion",
    color = "Diagnosis"
  )+  theme_minimal() + theme(legend.position = "none")
```


```{r DISCARD - efforts to make umap in bulk data}
all(dimnames(Z_case_mean)[[1]] == dimnames(Z_control_mean)[[1]])
genes <- dimnames(Z_case_mean)[[1]]
# Combine cell types
cell_types_case <- dimnames(Z_case_mean)[[2]]
cell_types_control <- dimnames(Z_control_mean)[[2]]

combined_cell_types <- c(cell_types_case, cell_types_control)

# Combine samples
samples_case <- dimnames(Z_case_mean)[[3]]
samples_control <- dimnames(Z_control_mean)[[3]]

combined_samples <- c(samples_case, samples_control)

# Initialize the combined array: Genes x Combined Cell Types x Combined Samples
combined_array <- array(NA, dim = c(length(genes), length(combined_cell_types), length(combined_samples)),
                        dimnames = list(genes, combined_cell_types, combined_samples))

# Fill in the combined array
# 1. Insert Z_case_mean
combined_array[, cell_types_case, samples_case] <- Z_case_mean

# 2. Insert Z_control_mean
combined_array[, cell_types_control, samples_control] <- Z_control_mean

# Check the dimensions of the combined array
dim(combined_array)

# View the combined array structure
str(combined_array)

# TRIED to do umap on the gene expression matrix but the UMAP was too weird, possibly due to the expression selective to cell types. 
Z_mean = combined_array

# Assuming Z_mean is a 3D array: Genes x CellTypes x Samples

# Step 1: Extract dimension names
genes <- dimnames(Z_mean)[[1]]        # Gene names
cell_types <- dimnames(Z_mean)[[2]]   # Cell type names
samples <- dimnames(Z_mean)[[3]]      # Sample names

# Step 2: Create combined "Sample_CellType" row identifiers
row_ids <- as.vector(outer(samples, cell_types, paste, sep = "_"))  # "Sample_CellType"

# Step 3: Initialize a matrix for the reshaped data
Z_flat <- matrix(NA, nrow = length(row_ids), ncol = length(genes))  # Initialize matrix
rownames(Z_flat) <- row_ids  # Assign row names (Sample_CellType)
colnames(Z_flat) <- genes     # Assign column names (genes)

# Step 4: Fill the flattened matrix
row_idx <- 1
for (sample in samples) {
  for (cell_type in cell_types) {
    Z_flat[row_idx, ] <- Z_mean[, cell_type, sample]
    row_idx <- row_idx + 1
  }
}

# Check the reshaped matrix
dim(Z_flat)  # Rows = Sample_CellType, Columns = Genes
colnames(Z_flat)[1:10]
# Step 1: Remove NAs and zeros
Z_flat_clean <- Z_flat  # Make a copy to keep the original intact

# Replace NAs and zeros with NA
Z_flat_clean[Z_flat_clean == 0] <- NA  # Treat 0 as missing
Z_flat_clean <- Z_flat_clean[rowSums(is.na(Z_flat_clean)) < ncol(Z_flat_clean), ]  # Drop rows with all NA
Z_flat_clean <- Z_flat_clean[, colSums(is.na(Z_flat_clean)) < nrow(Z_flat_clean)]  # Drop columns with all NA

# Replace remaining NAs with row or column means
Z_flat_clean[is.na(Z_flat_clean)] <- 0  # Replace NA with zero or use row/col means if needed
dim(Z_flat_clean)
Z_flat_clean <- scale(Z_flat_clean)  # Standardize the data
# Step 2: Calculate variance for genes (columns) and filter top 5000 genes
gene_variance <- apply(Z_flat_clean, 2, var, na.rm = TRUE)  # Variance for each gene
gene_variance <- sort(gene_variance, decreasing = TRUE)     # Sort by decreasing variance

top_5000_genes <- names(gene_variance[1:5000])  # Select top 5000 genes
Z_top_genes <- Z_flat_clean[, top_5000_genes, drop = FALSE]  # Subset the matrix

# Step 3: Calculate variance for samples (rows) and filter top 1000 samples
sample_variance <- apply(Z_top_genes, 1, var, na.rm = TRUE)  # Variance for each sample
sample_variance <- sort(sample_variance, decreasing = TRUE)  # Sort by decreasing variance

top_1000_samples <- names(sample_variance[1:1000])  # Select top 1000 samples
Z_top <- Z_top_genes[top_1000_samples, , drop = FALSE]  # Subset the matrix

# Step 4: Check the final dimensions
dim(Z_top)  # Should be 1000 rows (samples) and 5000 columns (genes)

# View a preview of the cleaned and filtered data
head(Z_top)

pca_result <- prcomp(Z_flat_clean, center = TRUE, scale. = TRUE)

# Step 3: Select the top 50 principal components
pca_top <- pca_result$x[, 1:50]  # Take the first 50 PCs

library(umap)
library(ggplot2)
umap_config <- umap.defaults
umap_config$n_neighbors <- 30  # Increase neighborhood size
umap_config$min_dist <- 0.2   
# Run UMAP on Z_flat
umap_result <- umap(pca_top)

# Prepare data for plotting
umap_df <- data.frame(
  UMAP1 = umap_result$layout[, 1],
  UMAP2 = umap_result$layout[, 2]
)

# Plot UMAP
ggplot(umap_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 0.5) +
 # geom_text(size = 2, vjust = 1.5) +
  labs(title = "UMAP of Sample-CellType Gene Expression", x = "UMAP1", y = "UMAP2") +
  theme_minimal()
```


## Risk prediction 

```{r}
# use dat obtained from merging cell props with patient data

# explore if cell type proportions differ by diagnosis
summary( glm (Astrocytes ~ Diagnosis + Age + Sex + PMI  , data = dat))
summary( glm (`Excitatory neurons` ~ Diagnosis + Age + Sex+ PMI, data = dat))
summary( glm (`Inhibitory neurons` ~ Diagnosis + Age + Sex+ PMI , data = dat))

# create finer categories for later modeling
dat$APOE_Risk <- factor(
  ifelse(dat$APoE %in% c("e43", "e42", "e44"), "High Risk", 
         ifelse(dat$APoE %in% c("e33", "e32"), "Low Risk", NA)))
dat$APOE_Risk <- factor(dat$APOE_Risk, levels = c("Low Risk", "High Risk"))

dat$Diagnosis <- ifelse(dat$Diagnosis == "AD", 1, 0)
dat$Diagnosis <- as.factor(dat$Diagnosis)

# remove rows with missing values in relevant columns and select columns for modeling
dat_clean <- na.omit(dat[, c("Diagnosis", "Astrocytes", "Excitatory neurons", "PMI", "Age", "Sex", "APOE_Risk")])

```


```{r}
# run risk prediction model and obtain performance evaluation metrics

# Model 1: Crude without adjustment for covariates
mod1 <- glm(Diagnosis ~ APOE_Risk, 
                      data = dat_clean, family = "binomial")

# Model 2: Adjusted for patient characteristics
mod2 <- glm(Diagnosis ~ Age + Sex + APOE_Risk, 
                      data = dat_clean, family = "binomial")

# Model 3: Unadjusted (excludes Astrocytes and Excitatory neurons)
mod3 <- glm(Diagnosis ~ Age + Sex + APOE_Risk + Astrocytes + `Excitatory neurons`, 
                        data = dat_clean, family = "binomial")

# Predicted probabilities
dat_clean$pred1 <- predict(mod1, type = "response")
dat_clean$pred2 <- predict(mod2, type = "response")
dat_clean$pred3 <- predict(mod3, type = "response")

library(pROC)

roc1 <- roc(dat_clean$Diagnosis, dat_clean$pred1)
auc1 <- auc(roc1)
# ROC curve and AUC for the adjusted model
roc2 <- roc(dat_clean$Diagnosis, dat_clean$pred2)
auc2 <- auc(roc2)

# ROC curve and AUC for the unadjusted model
roc3 <- roc(dat_clean$Diagnosis, dat_clean$pred3)
auc3 <- auc(roc3)

# Print AUC values
print(paste("AUC for model 1:", round(auc1, 3)))
print(paste("AUC for model 2:", round(auc2, 3)))
print(paste("AUC for model 3:", round(auc3, 3)))
      
# Plot ROC curves
plot(roc1, col = "blue", main = "ROC Curves for Risk Prediction Models")
plot(roc2, col = "red", add = TRUE)
plot(roc3, col = "green", add = TRUE)
legend("bottomright", legend = c("1", "2", "3"), col = c("blue", "red", "green"), lty = 1)

```


## Gene expression signature 

```{r}
# Load the output from the Gibbs sampler
res_case <- readRDS("./output/bulk_case_output.rds")
res_control <- readRDS("./output/bulk_control_output.rds")

# Obtain the mean gene expression values for each cell type
Z_case_mean <- apply(res_case$Z_n, c(1, 2, 3), mean)
Z_control_mean <- apply(res_control$Z_n, c(1, 2, 3), mean)

# Define patterns to search for
cell_type_patterns <- c("Astrocytes", "Excitatory neurons", "Inhibitory neurons")
# Apply the aggregation function from helper_functions.R
Z_case_agg <- aggregate_by_prefix(Z_case_mean, cell_type_patterns)
Z_control_agg <- aggregate_by_prefix(Z_control_mean, cell_type_patterns)

# Reshape aggregated Z_agg into a unified long-format data frame
library(reshape2)
Z_df_agg <- do.call(rbind, lapply(names(Z_agg), function(cell_type) {
  data.frame(
    Gene = rep(rownames(Z_agg[[cell_type]]), ncol(Z_agg[[cell_type]])),
    Sample = rep(colnames(Z_agg[[cell_type]]), each = nrow(Z_agg[[cell_type]])),
    CellType = cell_type,
    Expression = as.vector(Z_agg[[cell_type]])
  )
}))
Z_df_agg_case <- reshape_expression_list(Z_case_agg)
Z_df_agg_control <- reshape_expression_list(Z_control_agg)
Z_df_agg <- rbind(Z_df_agg_case, Z_df_agg_control)
# Inspect the reshaped data
tail(Z_df_agg)

## use previously calculated proportions
proportions_df = prop_summed
colnames(proportions_df)[1] <- "Sample"
# Convert proportions data frame to long format
proportions_long <- proportions_df %>%
  pivot_longer(cols = -Sample, names_to = "CellType", values_to = "Proportion")

# Merge aggregated gene expression with pre-calculated proportions
combined_data <- merge(Z_df_agg, proportions_long, by = c("Sample", "CellType"))

# Inspect combined data
head(combined_data)
```


```{r}
# Gene expression signature of AD-associated genes
library(dplyr)

# Option 1: Define a list of AD-associated genes
ad_genes <- c("APOE", "TREM2", "MAPT", "BIN1", "CLU", "GFAP", "CD33")

# Option 2: Load a pre-defined gene set from an external source
library(msigdbr)
library(dplyr)
library(psych)

# Retrieve KEGG gene sets for humans
kegg_human <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG")

# Filter for Alzheimer’s disease pathway
ad_kegg <- kegg_human %>% filter(gs_name == "KEGG_ALZHEIMERS_DISEASE")

# Filter for AD genes and compute correlations
# Filter the combined data for AD-related genes
ad_data <- combined_data %>% filter(Gene %in% ad_kegg$gene_symbol)

# Summarize average expression of AD gene signatures per sample and cell type
ad_signature_scores <- ad_data %>%
  group_by(Sample, CellType) %>%
  summarize(SignatureExpression = mean(Expression, na.rm = TRUE)) %>%
  merge(dat, by.x = "Sample", by.y="SampleID")

# Merge signature scores with cell-type proportions
merged_data <- merge(ad_signature_scores, proportions_long, by = c("Sample", "CellType"))

# Compute correlation between proportions and AD signature expression
cor_results <- merged_data %>%
  group_by(CellType) %>%
  summarize(
    Correlation = cor(Proportion, SignatureExpression, method = "pearson", use = "complete.obs"),
    p_value = corr.test(Proportion, SignatureExpression)$p.adj
  )

# Print correlation results
print(cor_results)

library(ggplot2)
library(viridis)

# Scatter plot for each cell type
p<-ggplot(ad_signature_scores %>% filter(CellType == "Astrocytes"), aes(x = Astrocytes, y = SignatureExpression, color = SignatureExpression)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey", alpha=0.25) +
  geom_point(size = 2, alpha=0.6) + scale_color_viridis() +
    annotate("text", x = 0.5, 
           y = 4.7, 
           label = paste0("Pearson's r = 0.74"),
           hjust = 0, vjust = 1, size = 4, color = "black") +
  ggtitle("Astrocytes") +
  xlab("Deconvoluted Proportion") +
  ylab("AD Signature Expression") +
  theme_minimal() + theme(legend.position = "none")
ggsave("./output/astrocytes_scatter.png", plot = p, width = 3, height = 4, dpi = 300)


p<-ggplot(ad_signature_scores %>% filter(CellType == "Excitatory neurons"), aes(x = `Excitatory neurons`, y = SignatureExpression, color = SignatureExpression)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey", alpha=0.25) +
  geom_point(size = 2, alpha=0.6) + scale_color_viridis() +
    annotate("text", x = 0.14, 
           y = 1.75, 
           label = paste0("Pearson's r = 0.96"),
           hjust = 0, vjust = 1, size = 4, color = "black") +
  ggtitle("Excitatory neurons") +
  xlab("Deconvoluted Proportion") +
  ylab("AD Signature Expression") +
  theme_minimal() + theme(legend.position = "none")
ggsave("./output/ex_scatter.png", plot = p, width = 3, height = 4, dpi = 300)

p<-ggplot(ad_signature_scores %>% filter(CellType == "Inhibitory neurons"), aes(x = `Inhibitory neurons`, y = SignatureExpression, color = SignatureExpression)) +
  geom_smooth(method = "lm", se = TRUE, color = "grey", alpha=0.25) +
  geom_point(size = 2, alpha=0.6) + scale_color_viridis() +
    annotate("text", x = 0.23, 
           y = 2.6, 
           label = paste0("Pearson's r = 0.82"),
           hjust = 0, vjust = 1, size = 4, color = "black") +
  ggtitle("Inhibitory neurons") +
  xlab("Deconvoluted Proportion") +
  ylab("AD Signature Expression") +
  theme_minimal() + theme(legend.position = "none")
ggsave("./output/in_scatter.png", plot = p, width = 3, height = 4, dpi = 300)
```


# additional 
```{r}
# try a different bulk 
bulk <- read.csv("./data/weighted_bulk_cpm.csv", row.names = 1)
bulk <- read.csv("./data/pseudobulk_cpm.csv", row.names = 1)
ref <- read.csv("./data/reference.csv", row.names = 1)

saveRDS(bulk, file = "./data/bulk.rds")
saveRDS(ref, file = "./data/ref.rds")

res <- readRDS("./output/bulk_pseudo_output.rds")
prop <- as.data.frame(res$theta_n)
prop_all <- process_sample_proportions(prop)
prop_summed <- prop_all %>%
  group_by(sample, cell_type) %>%
  summarise(total_proportion = sum(proportion, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(names_from = cell_type, values_from = total_proportion)

ggplot(prop_all, aes(x = cell_type, y = proportion, fill = reference)) +
  geom_bar(stat = "identity") + scale_fill_viridis(discrete = TRUE) +
  labs(
    title = "Stacked Bar Plot of Reference Contribution by Cell Type",
    x = "Cell Type",
    y = "Proportion",
    fill = "Reference"
  ) +
  theme_minimal()
```

